---
title: "Introduction to spatial tasks with R"
author: "Francois Rousseu"
date: '2017-07-17'
output:
  html_document:
    depth: 4
    number_sections: yes
    toc: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 4
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
geometry: left=2cm,right=2cm,top=3cm,bottom=2.5cm
---

<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"cef.gif\" style=\"float: right;width: 75px;\"/>')
   });
</script>

<style>
pre.r {
    background-color: #CCCCCC !important;
}
</style>

```{r options, include=TRUE, cache=FALSE, echo=FALSE, message=FALSE}
options(width=80)
knitr::opts_chunk$set(cache=TRUE,comment="##",echo=TRUE,message=FALSE,warning=FALSE,tidy=TRUE,size="small",fig.align="center")

```

# Introduction

The main book for learning to use R for spatial data is probably [Applied Spatial Data Analysis with R](http://www.asdar-book.org/) by Bivand et al. (2013)

## Packages

## Polygons

### locator
### getpoly

## `sp` vs. `sf` ?
  
# Shapefiles

## Reading and writing

Reading shaepfile is done with the function `readOGR` from the `rgdal` package
```{r read}
library(rgdal)
route<-readOGR("C:/Users/rouf1703/Documents/UdeS/Consultation/M-LLecuyer/Doc",layer="carreteras",verbose=FALSE)
class(route)
plot(route)

```

```{r write}

```

## Accessing data with S4 class

## The `Spatial` class with `sp`

```{r class}
library(sp)
getClass("Spatial")
```

## Building shapefiles

### From scratch

```{r shpscratch}

set.seed(123)
n<-10
x<-rnorm(n,0,1)
y<-rnorm(n,0,1)
id<-1:n
shp<-SpatialPointsDataFrame(cbind(x,y),data=data.frame(id))
head(shp)
plot(shp)

```


```{r shpscratchpol}

library(magrittr)

m<-rbind(c(1,1),c(1,2),c(2,2),c(2,1))
z<-rbind(m, m[1, ]) %>% Polygon %>% list %>% Polygons(ID = 1) %>% list %>% SpatialPolygons
plot(z)
```

### From a `data.frame`

```{r shpdf}
data(meuse)
head(meuse)
coordinates(meuse)<-~x+y
class(meuse)
plot(meuse)

```

## Spatial operations

Most spatial operation are done using the package `rgeos`. One of the requirement of the functions is that objects should be projected. When objects are really large, some operation may start to take a long time (e.g. intersections).

### Overlay things



### Cut or delimit things

Here are some common operations that can be done on points (but also lines or polygons.

```{r points_op, fig.width=7,fig.height=4,out.width="100%"}
library(rgeos)

op<-c("gBuffer","gConvexHull","gCentroid","gEnvelope","gDelaunayTriangulation")
par(mfrow=c(2,3),mar=c(0,0,3,0))
for(i in seq_along(op)){
  plot(shp,main=op[i],xlim=c(-2,2),ylim=c(-2,2))
  plot(get(op[i])(shp),col=gray(0.5,0.5),add=TRUE,pch=16,cex=3)
}

```

Here are some common opetrations that are done on polygons.

```{r polygons_op, fig.height=2}

x<-gBuffer(shp[1,],width=1)
y<-gBuffer(shp[2,],width=1)

op<-c("gIntersection","gDifference","gSymdifference","gUnion")
par(mfrow=c(2,2),mar=c(0,0,3,0))
for(i in seq_along(op)){
  plot(x,main=op[i],xlim=c(-2,2),ylim=c(-2,2))
  plot(y,add=TRUE)
  plot(get(op[i])(x,y),add=TRUE,col="red")
}

```

### Measure things

```{r measure_distance, fig.width=7,fig.height=3,out.width="100%"}

gDistance(shp,byid=TRUE)

```

```{r measure_area, fig.width=7,fig.height=5,out.width="100%"}

ch<-gDelaunayTriangulation(shp)
area<-gArea(ch,byid=TRUE)
plot(ch,axes=TRUE)
text(coordinates(ch)[,1],coordinates(ch)[,2],round(area,2),cex=0.7)

```

### Simplify things

```{r simplify, fig.width=3,fig.height=3,out.width="100%"}

n<-50
x<-rnorm(n)
y<-rnorm(n)
poly<-gBuffer(SpatialPoints(cbind(x,y)),width=0.5)
plot(poly)
plot(gSimplify(poly,tol=0.6,topologyPreserve=TRUE),border="red",add=TRUE)

```

#### spsample, elide, mst

## Plotting

### plot

```{r plot, fig.width=3,fig.height=3,out.width="100%"}

par(mar=c(0,0,0,0))
plot(ch,border="red",col="blue",lwd=3)

```

### spplot

```{r spplot, fig.width=6.5,fig.height=4,out.width="100%"}

spplot(meuse)

```

## The new package sf

# Projections

As always, when playing with spatial data, one needs to be aware of projections. Assigning and transforming projections in R is not to difficult. It is mostly done with the `proj4string` and `spTransform` functions from package **rgdal**. 

```{r proj_data, fig.width=6.5,fig.height=4,out.width="100%"}

library(rgbif)

x<-occ_search(scientificName="Phragmites australis",limit=500,hasCoordinate=TRUE,country="CA")
x<-as.data.frame(x$data)
coordinates(x)<-~decimalLongitude+decimalLatitude

```

## Assign a projection

```{r proj_assign}

proj4string(x)
proj4string(x)<-CRS("+proj=longlat +datum=WGS84 +ellps=WGS84")
#proj4string(x)<-CRS("+init=epsg:4326")
proj4string(x)

```

## Change a projection

```{r proj_change, fig.width=10, fig.height=6}

canada<-readOGR("C:/Users/rouf1703/Documents/UdeS/Formation/Canada",layer="Canada")
proj4string(canada)

crs<-c("+proj=longlat +datum=WGS84 +ellps=WGS84",
       "+proj=utm +zone=18 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0",
       "+proj=utm +zone=10 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0",
       "+proj=laea +lat_0=34 +lon_0=-10 +ellps=GRS80")

par(mfrow=c(2,2),mar=c(2,3,2,0))

for(i in 1:length(crs)){
  canada2<-spTransform(canada,CRS(crs[i]))
  x2<-spTransform(x,CRS(crs[i]))
  plot(canada2,axes=TRUE,main=crs[i],cex.main=0.7)
  plot(x2,add=TRUE) 
  
}


```

## Projections and EPSG

Projections can also be given with their epsg number. More info on projections can be found here [epsg.io](http://epsg.io/) where a list of the different epsg is available

```{r proj_epsg}

x<-spTransform(x,CRS("+init=epsg:4326"))
proj4string(x)

```

# Rasters

## What is a raster?

## Building a raster

Here is an example of building a simple raster with random values.
```{r buildrastersimple}
library(raster)

n<-50
r<-raster(nrow=50,ncol=50,ext=extent(canada))
r<-setValues(r,runif(n^2))
ncell(r)
plot(r)
plot(canada,add=TRUE)

```

Here is an example of building a raster where each pixel value is determined by the number of points in each cell.
```{r buildrastercomplex}
library(raster)
library(rgbif)

x<-occ_search(scientificName="Catharus bicknelli",limit=5000,hasCoordinate=TRUE,country="CA")
x<-as.data.frame(x$data)
coordinates(x)<-~decimalLongitude+decimalLatitude

r<-raster(nrow=50,ncol=50,ext=extent(canada))
r<-rasterize(x,r,field=1,fun="count",background=0)

plot(r,col=rev(heat.colors(100)))
plot(canada,col=gray(0,0.05),border=gray(0,0.01),add=TRUE)
points(x,col=gray(0,0.1))

```

## Raster formats

## Read, write, merge, crop

```{r read_rast}

r1<-raster("C:/Users/rouf1703/Documents/UdeS/Formation/cdem_dem_021E_tif/cdem_dem_021E.tif")
r2<-raster("C:/Users/rouf1703/Documents/UdeS/Formation/cdem_dem_031H_tif/cdem_dem_031H.tif")
par(mfrow=1:2)
plot(r1)
plot(r2)

r1

```


```{r merge}

r1<-aggregate(r1,10)
r2<-aggregate(r2,10)
r<-merge(r1,r2)
plot(r)

```


```{r crop}

e <- extent(-72.75, -70.25, 45, 46)
r <- crop(r, e)
plot(r)

```

## Stacks and Bricks

```{r stack}



```


## Raster operations

### Summarize


```{r summ}

r

```

### Extract

```{r extract}

e<-extract(r,x)
hist(e,xlab="Altitude",main="")

```

### Reclassify

```{r reclassify}

mn<-c(0,100,200,400,600,800)
mx<-c(100,200,400,600,800,1200)
mat<-cbind(mn,mx,lab=mx)
mat
rc<-reclassify(r,mat)
plot(rc,col=terrain.colors(nrow(mat)))

```


### Countour and rasterToContours

### Speeding things up with `velox`

```{r velox}
library(velox)
library(microbenchmark)

rv<-velox(r)

buff<-gBuffer(x,byid=TRUE,width=0.02)
plot(r)
plot(buff,add=TRUE)

system.time(extract(r,buff,fun=mean))
system.time(rv$extract(buff,fun=mean))

```

## From shp to raster and back

### Rasterizing a polygon

```{r rasterizepol}

pr<-rasterize(buff,r)
pr
plot(pr)

```

### Polygonizing a raster

## Plotting with `rasterVis`

# `SpatialGrids` and `SpatialPixels`

## Grids

## Pixels

# Static Mapping

## `plot`

## `levelplot`

## `spplot`

## `ggplot` and `ggmap`

## `ggspatial`

## chloropleth mapping

# Interactive Mapping

## `OpenStreetMap`

## `leaflet`

## `tmap` 

## `mapview`

## `mapedit`

Here is a test with using mapedit functionality.

```{r mapedit, eval=FALSE}
library(mapview)
library(mapedit)
library(sp)

x <- mapview() %>% editMap()
plot(x)
x <- as(x[[1]],"Spatial")
plot(x)

```

# More possibilities

## Spatial Task View

## Reference Site: rspatial.org

## Special packages

### Spatial statistics with `spatstat`

### Spatial dependence with `spdep`

### Kriging with `gstat`

### Animal movements with `adehabitat` and `trip`

### Kernel smoothing with `ks`

### Great circles with `geosphere`

## Extensions

### RQGIS

### RpostGIS

### RGRASS

# `sf`: The new spatial

## Why?

### S3 classes

### Tidyverse

### From `sp` to `sf` and back

